Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-06 paper/main.tex
Updates

\section{Background and Related Work}
\label{sec:background}
\subsection{Notations}
We denote random variables with uppercase letters $ X $, and a set of random
variables as $ \bm{X} = \{X_1, \cdots, X_k\} $ with $ \rvert \bm{X} \rvert = k
$. A sample from the random variable $ X $ is denoted as $ x $ and from a set
of random variables $ \bm{X} $ is denoted as $ \bm{x} $. In this paper, we
consider random variables in the mixed data setting where each of the random
variables can be either of continuous, categorical, or ordinal unless
specified. We write the expectation of a variable $ X $ as $ \mathbb{E}[X] $,
conditional expectations as $ \mathbb{E}[X | \bm{Z}] $. We denote a graph $ G =
(V, E) $ which nodes $ V$ and edges $ E $. In this paper we focus on all
observed variables, and linear measure of association. 
where $ \mathrm{cov}(X, Y) $ is the covariance between $ X $ and $ Y $, and $
\sigma_X $ and $ \sigma_Y $ are the standard deviations of $ X $ and $ Y $
respectively.

@@

In the previous section, we discussed how different types of variables require
different conditional measures of association. However, no unified measure
exists for mixed data. Here, we introduce a measure of conditional association
for mixed data by extending the concept of partial correlation coefficient for
continuous variables to mixed data. Specifically, our approach combines a mixed
data residualization method \citep{Ankan2023} with Pillai's Trace
\citep{Pillai1955}, a multivariate measure of association based on canonical
correlations.

Given a dataset $ D = (x, y, \bm{z}) $ on variables $ X $, $ Y $, and $ \bm{Z}
$, we goal to estimate the conditional association $ \rho(X, Y; \bm{Z}) $ . We
achieve this by first computing the residuals $ R_X $ and $ R_Y $ for $ X $ 
and $ Y $ respectively, using the following residualization approach:

@@

Here, we have the option to choose the estimators based on the properties
(distribution, type of relationship, and so on) of the data. Non-parametric
ensemble estimators such as Random Forest and XGBoost can handle any
combination of data types while performing reasonably well on a wide range of
distributions and relationship types. For linear relationships, we can also
choose simpler models such as linear regression and its variants.

Repeating the above residualization process for both $ X $ and  $ Y $, we
obtain residual matrices $ R_X $ and $ R_Y $. The type of the variable
determines the shape for these matrices. If the variable is continuous or
ordinal, the residual matrix is of shape $ ( n \times 1 ) $ and if the variable
is categorical, we have a residual matrix of shape $ ( x \times (k-1)) $ where
$ k $ is the number of categories of the variable.

The next step is to quantify the association between these residual matrices.
For this purpose we use canonical correlations that have been widely used to
measure the association between sets of random variables.

\begin{definition}

	Given two sets of random variables $ \bm{U} = (U_1, U_2, \cdots, U_p) $
	and $ \bm{V} = (V_1, V_2, \cdots, V_q) $, canonical correlation between
	them, $\rho(\bm{U}, \bm{V}) $ is defined as:

@@

Canonical correlations generalize correlation coefficients to multi-dimensional
variables. They find orthogonal linear transformations $ a $ and $ b $ of $
\bm{U} $ and $ \bm{V} $ such that the correlation coefficient between the
transformed variables $ a^T \bm{U} $ and $ b^T \bm{V} $ is maximized. This
yields a vector of correlation coefficient values of size $ \min(p, q) $,
representing the correlation coefficient of each pair of transformed variables.
Pearson's correlation coefficient is a special case of canonical correlations
when $ p = q = 1 $.

Many measures of association based on canonical correlation have been used in
multivariate statistics, each with slightly different properties.
\begin{itemize}
	\item Wilks' Lambda: $\Lambda = \prod_{i}^{\min(p, q)} (1 - \phi_i^2) $
	\item Roy's Largest Root: $ \theta = \max_i(\phi_i^2) $
	\item Pillai's Trace: $ \tau = \sum_{i=1}^{\min(p, q)} \phi_i^2 $
\end{itemize}

We use a slightly modified version of Pillai's Trace for our purpose for two
main reasons: 1) It uses all the canonical correlation values, 2) Its
interpretation is similar to correlation coefficient, i.e., $ 0 $ signifies no
association and $ 1 $ signifies perfect linear relationship. This gives us the
following partial measure of association:

@@

Our measure of conditional association has several key properties that make it
well-suited for our application:

\begin{enumerate}
	\item \textbf{Bounded: } $ \rho $ is bounded between $ 0 $ (no
		association) to $ 1 $ (perfect linear association). This bounded
		nature simplifies the interpretation and allows for easy 
		comparisons across multiple values.
	\item \textbf{Independent of Sample Size: } 
	\item \textbf{Invariant to the Dummy Encoding: } We use dummy encoding 
		to compute the residual matrix for categorical variables. Since
		canonical correlations identify linear combinations across
		columns, this measure is invariant to the specific dummy
		encoding scheme used.
	\item \textbf{Equivalent to Partial Correlation}
		When both $ X $ and $ Y $ are continuous variables, our measure
		is equivalent to the absolute value of partial correlation.
	\item \textbf{Equivalent to polychoric and polyserial correlation for
			ordinal and continuous variables: }
		Under the assumption that the ordinal variable is generated by
		discretizing an underlying gaussian continuous variable, this
		effect size is equivalent to polychoric and polyserial
		correlation. As both of them recover the Pearson correlation
		coefficient.
	\item \textbf{Related to Cram\'er's V: } 
\end{enumerate}

So, in a sense this measure generalized currently used measures of associations
to mixed data.

>>>>

\section{Background and Related Work}
\label{sec:background}
\subsection{Notations} 
\todo[inline]{Needs improvement}
We denote random variables with uppercase letters $ X $, and a set of random
variables as $ \bm{X} = \{X_1, \cdots, X_k\} $ with $ \rvert \bm{X} \rvert = k
$. A sample from the random variable $ X $ is denoted as $ x $ and from a set
of random variables $ \bm{X} $ is denoted as $ \bm{x} $. In this paper, we
consider random variables in the mixed data setting where each of the random
variables can be either of continuous, categorical, or ordinal unless
specified. We write the expectation of a variable $ X $ as $ \mathbb{E}[X] $,
conditional expectations as $ \mathbb{E}[X | \bm{Z}] $. We denote a graph $ G =
(V, E) $ which nodes $ V$ and edges $ E $. In this paper we focus on all
observed variables, and linear measure of association. 
where $ \mathrm{cov}(X, Y) $ is the covariance between $ X $ and $ Y $, and $
\sigma_X $ and $ \sigma_Y $ are the standard deviations of $ X $ and $ Y $
respectively.

@@

In the previous section, we discussed how different types of variables require
different measures of conditional association. However, there is no unified
measure that handles mixed data. Here, we introduce a measure of conditional
association for mixed data by extending the concept of partial correlation
coefficient (commonly used for continuous variables) to mixed data. Similar to
partial correlation coefficient, our method integrates a mixed data
residualization method \citep{Ankan2023} with Pillai's Trace
\citep{Pillai1955}, a multivariate measure of association based on canonical
correlations.

Given a dataset $ D = (x, y, \bm{z}) $ on variables $ X $, $ Y $, and $ \bm{Z}
$, our goal is to estimate the conditional association $ \rho(X, Y; \bm{Z}) $. 
In the first step, we compute the residuals $ R_X $ and $ R_Y $ for variables
$ X $ and $ Y $ respectively. Depending on the type of variable the residual
is computed as follows:

@@

We have the option here to choose the estimators based on the characteristics
such as distribution, type of relationship, and so on of our dataset.
Non-parametric ensemble estimators such as Random Forest and XGBoost are robust 
for diverse data types and complex relationships. For linear relationships,
simpler models such as linear regression and its variants may suffice.

We repeat the above residualization step for both the variables $ X $ and $ Y $
obtaining residual matrices $ R_X $ and $ R_Y $. The type of variable determines
the shape of these matrices. If the variable is continuous or ordinal, the 
residual matrix is of shape $ (n \times 1 ) $ and if the variable is categorical,
we get a residual matrix of shape $ (x \times (k - 1)) $, where $ k $ is the number
of categories of the variable.

The second step is to quantify the association between these residual matrices.
For this purpose we use canonical correlations that have been widely used to
measure the association between sets of random variables.

\begin{definition}
	Given two sets of random variables $ \bm{U} = (U_1, U_2, \cdots, U_p) $
	and $ \bm{V} = (V_1, V_2, \cdots, V_q) $, canonical correlation between
	them, $\rho(\bm{U}, \bm{V}) $ is defined as:

@@

Canonical correlations generalize the concept of correlation coefficients to
multi-dimensional variables. It finds orthogonal linear transformations $ a $
and $ b $ that maximized the correlation between the transformed variables $
a^T \bm{U} $ and $ b^T \bm{V} $. This yields a vector of correlation
coefficient values of size $ \min(p, q) $, representing the correlation
coefficient of each pair of transformed variables. Notably, Pearson's
correlation coefficient is a special case of canonical correlations when $ p =
q = 1 $.

Several measures of association have been derived from canonical correlations, such as:
\begin{itemize}
	\item Wilks' Lambda: $\Lambda = \prod_{i}^{\min(p, q)} (1 - \phi_i^2) $
	\item Roy's Largest Root: $ \theta = \max_i(\phi_i^2) $
	\item Pillai's Trace: $ \tau = \sum_{i=1}^{\min(p, q)} \phi_i^2 $
\end{itemize}

We use a slightly modified version of Pillai's Trace for our purpose for two
main reasons: 1) It uses all the canonical correlation values, capturing the
full extent of association, 2) Its interpretation is similar to Pearson's
correlation coefficient, i.e., $ 0 $ signifies no association and $ 1 $
signifies perfect linear relationship. This leads to our proposed measure of
conditional association

@@

Our proposed measure has several desirable properties that make it well-suited
for our application:

\begin{enumerate}
	\item \textbf{Bounded: } The measure is bounded between $ 0 $ (no
		association) to $ 1 $ (perfect linear association), simplifying
		interpretation.
	\item \textbf{Independent of Sample Size: } 
	\item \textbf{Invariant to the Dummy Encoding: } For categorical
		variables, residuals are computed using dummy encoding. Since
		canonical correlations identifies linear combinations across
		columns, this measure is invariant to the specific dummy
		encoding scheme used.
	\item \textbf{Equivalent to Partial Correlation for Continuous Variables: }
		When both $ X $ and $ Y $ are continuous variables, our measure
		is equivalent to the absolute value of partial correlation coefficient.
	\item \textbf{Equivalent to polychoric and polyserial correlation for
			ordinal and continuous variables: }
		Under the assumption that the ordinal variable is generated by
		discretizing an underlying gaussian continuous variable, this
		effect size is equivalent to polychoric and polyserial
		correlation. As both of them recover the Pearson correlation
		coefficient.
	\item \textbf{Related to Cram\'er's V: } 
\end{enumerate}

In essence, our measure of conditional association extends existing metrics to mixed data, providing 
a unified measure that is interpretable.
